---
author: 'Catherine Greene'
title: 'Bios 6301: Assignment 7'
output: pdf_document
---

*Due Thursday, 03 November, 1:00 PM*

$5^{n=day}$ points taken off for each day late.

40 points total.

Submit a single knitr file (named `homework7.rmd`), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as `author` to the file's metadata section. Raw R code/output or word processor files are not acceptable.

Failure to name file `homework7.rmd` or include author name may result in 5 points taken off.

### Question 1 ###

**21 points**

Use the following code to generate data for patients with repeated measures of A1C (a test for levels of blood glucose).

```{r}
genData <- function(n) {
    if(exists(".Random.seed", envir = .GlobalEnv)) {
        save.seed <- get(".Random.seed", envir= .GlobalEnv)
        on.exit(assign(".Random.seed", save.seed, envir = .GlobalEnv))
    } else {
        on.exit(rm(".Random.seed", envir = .GlobalEnv))
    }
    set.seed(n)
    subj <- ceiling(n / 10)
    id <- sample(subj, n, replace=TRUE)
    times <- as.integer(difftime(as.POSIXct("2005-01-01"), as.POSIXct("2000-01-01"), units='secs'))
    dt <- as.POSIXct(sample(times, n), origin='2000-01-01')
    mu <- runif(subj, 4, 10)
    a1c <- unsplit(mapply(rnorm, tabulate(id), mu, SIMPLIFY=FALSE), id)
    data.frame(id, dt, a1c)
}
x <- genData(500)
```

Perform the following manipulations: (3 points each)

1. Order the data set by `id` and `dt`.
```{r}
#table(x['id']) # each id has multiple instances, sort by dt within each id group
x_ordered <- x[order(x[,'id'], x['dt']),]
rownames(x_ordered) <- 1:nrow(x_ordered) # reset the indices
```
2. For each `id`, determine if there is more than a one year gap in between observations.  Add a new row at the one year mark, with the `a1c` value set to missing.  A two year gap would require two new rows, and so forth.
```{r echo=TRUE}
install.packages("lubridate", repos = "http://cran.us.r-project.org") # install lubridate
library("lubridate")

rows <- nrow(x_ordered)-1 # I was having trouble getting this to work inline

for (i in (1:rows)) {
  if (x_ordered[,'id'][i] == x_ordered[,'id'][i+1]) { # for the same id
    gap <-abs(difftime(x_ordered[,'dt'][i], x_ordered[,'dt'][i+1], units='days')) # what is the gap
  }
  if ((gap>365) && (gap<(2*365))) { # is the gap bigger than a year but less than 2 years
    gap_df <- data.frame(x_ordered[,'id'][i], x_ordered[,'dt'][i] %m+% years(1), 'NA')
    names(gap_df) <- c('id','dt','a1c')
    x_ordered <- rbind(x_ordered, gap_df)
    #print(gap)
  } else if (gap>(2*365)) { # is the gap bigger than 2 years
    gap_df1 <- data.frame(x_ordered[,'id'][i], x_ordered[,'dt'][i] %m+% years(1), 'NA')
    gap_df2 <- data.frame(x_ordered[,'id'][i], x_ordered[,'dt'][i] %m+% years(2), 'NA')
    names(gap_df1) <- c('id','dt','a1c')
    names(gap_df2) <- c('id','dt','a1c')
    x_ordered <- rbind(x_ordered, gap_df1, gap_df2) # I did it this way because it defaulted to wrong organization of rows vs columns
    #print (gap)
  }
}

x_ordered_update <- x_ordered[order(x_ordered[,'id'], x_ordered['dt']),]
rownames(x_ordered_update) <- 1:nrow(x_ordered) # reset the indices

```
3. Create a new column `visit`.  For each `id`, add the visit number.  This should be 1 to `n` where `n` is the number of observations for an individual.  This should include the observations created with missing a1c values.
```{r}
visit <- c()

for (i in 1:length(unique(x_ordered_update[,'id']))) {
  visit_sub <- c(seq(table(x_ordered_update[,'id'])[i]))
  visit <- c(visit, visit_sub)
}

x_visit <- cbind(x_ordered_update, visit)
```
4. For each `id`, replace missing values with the mean `a1c` value for that individual.
```{r}
# I know that each id is 1-50, so I'm going to make my life easier and just say that
a1c_update <- c()
for (i in 1:50) {
  subset <- suppressWarnings(as.numeric(x_visit[which(x_visit['id']==i),][['a1c']]))
  subset[which(is.na(subset) == TRUE)] <- mean(subset, na.rm = TRUE)
  a1c_update <- c(a1c_update, subset)
}

x_final <- cbind(x_visit, a1c_update)[,-3]
```
5. Print mean `a1c` for each `id`.
```{r}
# same as above, no NAs this time though
a1c_means <- c()

for (i in 1:50) {
  subset <- x_final[which(x_final['id']==i),][['a1c_update']]
  a1c_means[i] <- mean(subset)
}

a1c_means
```
6. Print total number of visits for each `id`.
```{r}
table(x_final['id'])
```
7. Print the observations for `id = 15`.
```{r}
x_final[which(x_final['id']==15),]
```

### Question 2 ###

**16 points**

Install the `lexicon` package. Load the `sw_fry_1000` vector, which contains 1,000 common words.

```{r}
install.packages('lexicon', repos = "http://cran.us.r-project.org")
data('sw_fry_1000', package = 'lexicon')
head(sw_fry_1000)
```

1. Remove all non-alphabetical characters and make all characters lowercase. Save the result as `a`.

```{r}
a <- tolower(grep("[^1-9]", sw_fry_1000, value=TRUE))
```

Use vector `a` for the following questions. (2 points each)

2. How many words contain the string "ar"?

```{r}
length(grep("ar", sw_fry_1000, value=TRUE))
```

3. Find a six-letter word that starts with "l" and ends with "r".

```{r}
grep("^l.*r${6}", sw_fry_1000, value=TRUE)
```

4. Return all words that start with "col" or end with "eck".

```{r}
grep("^col|eck$", sw_fry_1000, value=TRUE)
```

5. Find the number of words that contain 4 or more adjacent consonants. Assume "y" is always a consonant.

```{r}
grep("[b,c,d,f,g,h,j,k,l,m,n,p,q,r,s,t,v,w,x,y,z]{4}", sw_fry_1000, value=TRUE)
```

6. Return all words with a "q" that isn't followed by a "ui".

```{r}
grep("q.[^ui]", sw_fry_1000, value=TRUE)
```

7. Find all words that contain a "k" followed by another letter. Run the `table` command on the first character following the first "k" of each word.

```{r}
k <- grep("k[a-z]", sw_fry_1000, value=TRUE)
k_cut <- sub(".*k.*([a-z].*$)", "\\1", k)
table(k_cut)
```

8. Remove all vowels. How many character strings are found exactly once?

```{r}
no_vowels <- gsub("[aeiou]", "", sw_fry_1000)
length(unique(no_vowels)) #743 unique character strings, once vowels are cut out
```

### Question 3 ###

**3 points**

The first argument to most functions that fit linear models are formulas.  The following example defines the response variable `death` and allows the model to incorporate all other variables as terms. `.` is used to mean all columns not otherwise in the formula.

```{r}
#url <- "https://github.com/couthcommander/Bios6301/raw/master/datasets/haart.csv"
#haart_df <- read.csv(url)[,c('death','weight','hemoglobin','cd4baseline')]
#coef(summary(glm(death ~ ., data=haart_df, family=binomial(logit))))
```

Now imagine running the above several times, but with a different response and data set each time.  Here's a function:

```{r}
myfun <- function(dat, response) {
  form <- as.formula(response ~ .)
  coef(summary(glm(form, data=dat, family=binomial(logit))))
}
```

Unfortunately, it doesn't work. `tryCatch` is "catching" the error so that this file can be knit to PDF.

```{r}
tryCatch(myfun(haart_df, death), error = function(e) e)
```

What do you think is going on?  Consider using `debug` to trace the problem.

```{r}
#debug(myfun)
#myfun(haart_df, death)

# This may just be a problem on my end, but it seems that the issue is originating with the url. The web address may be faulty, which is not allowing the dataset to load.
```

**5 bonus points**

Create a working function.
