---
title: 'Bios 6301: Assignment 3'
author: "Catherine Greene"
output:
  pdf_document: default
  html_document: default
---

*Due Tuesday, 27 September, 1:00 PM*

50 points total.

Add your name as `author` to the file's metadata section.

Submit a single knitr file (named `homework3.rmd`) by email to tianyi.sun@vanderbilt.edu.
Place your R code in between the appropriate chunks for each question.
Check your output by using the `Knit HTML` button in RStudio.

$5^{n=day}$ points taken off for each day late.

### Question 1 ###

**15 points**

Write a simulation to calculate the power for the following study
design.  The study has two variables, treatment group and outcome.
There are two treatment groups (0, 1) and they should be assigned
randomly with equal probability.  The outcome should be a random normal
variable with a mean of 60 and standard deviation of 20.  If a patient
is in the treatment group, add 5 to the outcome.  5 is the true
treatment effect.  Create a linear model for the outcome by the
treatment group, and extract the p-value (hint: see assignment1).
Test if the p-value is less than or equal to the alpha level, which
should be set to 0.05.

Repeat this procedure 1000 times. The power is calculated by finding
the percentage of times the p-value is less than or equal to the alpha
level.  Use the `set.seed` command so that the professor can reproduce
your results.

```{r}
# scratchwork, I'm testing the code here

n <- 100 
newoutcome <- c()
set.seed(n)
treatment <- rbinom(n, 1, 0.5)
outcome <- rnorm(n, mean=60, sd=20)
for (i in 1:n) {
  if (treatment[i]==1){
    newoutcome[i] <- outcome[i] + 5}
  else
    newoutcome[i] <- outcome[i]
  }
  t.test(newoutcome ~ treatment, alternative='two.sided', mu=0)$p.value
# p-value = 0.2832892

# check that the treatment effect works
my_data <- data.frame(treatment, outcome, newoutcome)

# linear model
model <- lm(newoutcome ~ treatment)
get_p <- summary(model)$coefficients[2, 4] # extract the p-value from the linear model

```

1. Find the power when the sample size is 100 patients. (10 points)

```{r}
set.seed(100)

mean(replicate(1000, {
  treatment <- rbinom(100, 1, 0.5)
  outcome <- rnorm(100, mean=60, sd=20)
  for (i in 1:100) {
    if (treatment[i]==1){
      outcome[i] <- outcome[i] + 5}
    }
  t.test(outcome ~ treatment, alternative='two.sided', mu=0)$p.value
  }) < 0.05)

# power = 23.3%

```

1. Find the power when the sample size is 1000 patients. (5 points)

```{r}

set.seed(1000)

# how do I get the t.test working?
mean(replicate(1000, {
  treatment <- rbinom(1000, 1, 0.5)
  outcome <- rnorm(1000, mean=60, sd=20)
  for (i in 1:1000) {
    if (treatment[i]==1){
      outcome[i] <- outcome[i] + 5}
    }
  t.test(outcome ~ treatment, alternative='two.sided', mu=0)$p.value
  }) < 0.05)

# power = 96.8%
```

### Question 2 ###

**14 points**

Obtain a copy of the [football-values lecture](https://github.com/couthcommander/football-values).
Save the `2021/proj_wr21.csv` file in your working directory.  Read
in the data set and remove the first two columns.

```{r}
hw3_football <- read.csv("/Users/KatietheWise/Desktop/2022_Fall/StatComp/Homework/proj_wr21.csv")
hw3_football[, 'PlayerName'] <- NULL
hw3_football[, 'Team'] <- NULL

#summary(hw3_football)
```

1. Show the correlation matrix of this data set. (4 points)
```{r}
# grab some key aspects of the data set
means.football <- colMeans(hw3_football)
var.football <- var(hw3_football)
# correlation matrix
cor(hw3_football)

```

1. Generate a data set with 30 rows that has a similar correlation
structure.  Repeat the procedure 1,000 times and return the mean
correlation matrix. (10 points)

```{r}
# install.packages("MASS")
library(MASS)

# make the simulated data set
football.sim <- mvrnorm(30, mu = means.football, Sigma = var.football) # simulate from a multivariate normal distribution
football.sim <- as.data.frame(football.sim) # turn it into a data.frame
football.sim # show data.frame

# then loop it to figure out the average matrix
final_football <- 0 # start at 0
for (i in 1:1000){
  football.sim <- mvrnorm(30, mu = means.football, Sigma = var.football) # same as above
  final_football <- final_football + cor(football.sim)/1000
}
final_football # mean correlation matrix
```

### Question 3 ###

**21 points**

Here's some code:

```{r}
nDist <- function(n = 100) {
    df <- 10
    prob <- 1/3
    shape <- 1
    size <- 16
    list(
        beta = rbeta(n, shape1 = 5, shape2 = 45),
        binomial = rbinom(n, size, prob),
        chisquared = rchisq(n, df),
        exponential = rexp(n),
        f = rf(n, df1 = 11, df2 = 17),
        gamma = rgamma(n, shape),
        geometric = rgeom(n, prob),
        hypergeometric = rhyper(n, m = 50, n = 100, k = 8),
        lognormal = rlnorm(n),
        negbinomial = rnbinom(n, size, prob),
        normal = rnorm(n),
        poisson = rpois(n, lambda = 25),
        t = rt(n, df),
        uniform = runif(n),
        weibull = rweibull(n, shape)
    )
}
```

1. What does this do? (3 points)

    ```{r}
    round(sapply(nDist(500), mean), 2)
    ```
    
    ```
The expression above randomly samples 500 values (instead of the default 100 listed) from each of the given distributions, then calculates the mean for each distribution (`sapply()` applies the mean function to each set of 500). The means are then rounded to return values with 2 decimal points.
    ```

2. What about this? (3 points)

    ```{r}
    sort(apply(replicate(20, round(sapply(nDist(10000), mean), 2)), 1, sd))
    ```
    
    ```
The expression above randomly samples 10000 values from each of the given distributions, then calculates the mean for each distribution. The means are then rounded to return values with 2 decimal points. This is the same as the previous expression (#1). This procedure is repeated 20 times. The standard deviation for each distribution is then computed (each row), with the final result being distributions sorted in ascending order according to standard deviation.

    ```

    In the output above, a small value would indicate that `N=10,000` would provide a sufficent sample size as to estimate the mean of the distribution. Let's say that a value *less than 0.02* is "close enough".

1. For each distribution, estimate the sample size required to simulate the distribution's mean. (15 points)

Don't worry about being exact. It should already be clear that N < 10,000 for many of the distributions. You don't have to show your work. Put your answer to the right of the vertical bars (`|`) below.

    I just used approximate round numbers.

distribution|N
---|---
beta|10
binomial|10000
chisquared|60000
exponential|3000
f|2000
gamma|4000
geometric|11000
hypergeometric|5000
lognormal|10000
negbinomial|250000
normal|3000
poisson|70000
t|6000
uniform|300
weibull|3000
