---
author: 'Catherine Greene'
title: 'Bios 6301: Assignment 8'
output: pdf_document
---

*Due Tuesday, 15 November, 1:00 PM*

$5^{n=day}$ points taken off for each day late.

30 points total.

**30/30, great work! See comments for Q2. Tianyi**

Submit a single knitr file (named `homework8.rmd`), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as `author` to the file's metadata section. Raw R code/output or word processor files are not acceptable.

Failure to name file `homework8.rmd` or include author name may result in 5 points taken off.

### Question 1 ###

**15 points**

Install the `readxl` package and run the following

```{r echo=TRUE, error=TRUE}
install.packages("readxl", repos="http://cran.us.r-project.org") # it doesn't like my mirror
library(readxl)

fn <- 'icd10.xlsx'
if(file.access(fn, mode = 4) == -1) {
    url <- "https://www.cdc.gov/nhsn/xls/icd10-pcs-pcm-nhsn-opc.xlsx"
    download.file(url, destfile = fn, mode = 'wb')
}
dat <- readxl::read_excel(fn, sheet = 2)
```

1. Show the class of `dat`. (1 point)
```{r echo=TRUE, error=TRUE}
class(dat)
```

2. Show the methods available for objects of the given class (if there are multiple classes, show methods for all classes). (3 points)
```{r echo=TRUE, error=TRUE}
methods(class='tbl_df')
methods(class='tbl')
methods(class='data.frame')
```

3. If you call `print(dat)`, what print method is being dispatched? (1 point)
```{r echo=TRUE, error=TRUE}
#print(dat) # when I included this my document was ~500 pgs
# tbl_df
# uses tbl_df because that class is listed first in the vector when I searched for class(dat), and it has the print function listed as a method
```

4. Set the class of `dat` to be a data.frame. (1 point)
```{r echo=TRUE, error=TRUE}
class(dat) <- "data.frame"
```

5. If you call `print(dat)` again, what print method is being dispatched? (1 point)
```{r echo=TRUE, error=TRUE}
class(dat)
#print(dat)
# print.data.frame is the method applied since it's just a data.frame object now
```

Define a new generic function `nUnique` with the code below.

```{r echo=TRUE, error=TRUE}
nUnique <- function(x) {
    UseMethod('nUnique')
}
```

6. Write a default method for `nUnique` to count the number of unique values in an element. (2 points)
```{r echo=TRUE, error=TRUE}
nUnique.default <- function(x) {
  length(unique(x))
}
```

7. Check your function (2 points)

```{r, eval = FALSE}
nUnique(letters) # should return 26
nUnique(sample(10, 100, replace = TRUE)) # should return 10 (probably)
```

8. Write a data.frame method for `nUnique` to operate on data.frame objects.
This version should return counts for each column in a data.frame. (2 points)
```{r echo=TRUE, error=TRUE}
nUnique.data.frame <- function(x) {
  counts <- c()
  for (i in seq(ncol(x))) {
    counts <- c(counts, length(unique(x[,i])))
  }
  counts
}
```

9. Check your function (2 points)

```{r, eval = FALSE}
nUnique(dat)
```

### Question 2 ###

**15 points**

Programming with classes.  The following function will generate random patient information.

```{r}
makePatient <- function() {
  vowel <- grep("[aeiou]", letters)
  cons <- grep("[^aeiou]", letters)
  name <- paste(sample(LETTERS[cons], 1), sample(letters[vowel], 1), sample(letters[cons], 1), sep='')
  gender <- factor(sample(0:1, 1), levels=0:1, labels=c('female','male'))
  dob <- as.Date(sample(7500, 1), origin="1970-01-01")
  n <- sample(6, 1)
  doa <- as.Date(sample(1500, n), origin="2010-01-01")
  pulse <- round(rnorm(n, 80, 10))
  temp <- round(rnorm(n, 98.4, 0.3), 2)
  fluid <- round(runif(n), 2)
  list(name, gender, dob, doa, pulse, temp, fluid)
}
```

1. Create an S3 class `medicalRecord` for objects that are a list with the named elements `name`, `gender`, `date_of_birth`, `date_of_admission`, `pulse`, `temperature`, `fluid_intake`. Note that an individual patient may have multiple measurements for some measurements.  Set the RNG seed to `8` and create a medical record by taking the output of `makePatient`.  Print the medical record, and print the class of the medical record. (5 points)
```{r echo=TRUE, error=TRUE}
set.seed(8)
p <- makePatient() # I think this is what the instructions were saying, but I'd like to be able to learn how to make this more broadly applicable

patient_record <- list(name='Yes', gender='male', date_of_birth='1977-05-03', date_of_admission=c('2013-06-09', '2013-07-02'), pulse=c(79, 78), temperature=c(98.07, 97.50), fluid_intake=c(0.28, 0.52))
# TS: we would avoid manually enter these information
patient_record <- p
names(patient_record) = c('name','gender','date_of_birth','date_of_admission',
                'pulse','temperature','fluid_intake')

class(patient_record) <- 'medicalRecord'
attributes(patient_record)

print.medicalRecord <- function(record) {
  cat(sprintf("name: %s\ngender: %s\ndate_of_birth: %s\ndate_of_admission: %s\npulse: %s\ntemperature: %s\nfluid_intake: %s", 
              record$name, record$gender, record$date_of_birth, paste(record$date_of_admission, collapse=" "), paste(record$pulse, collapse=" "), paste(record$temperature, collapse=" "), paste(record$fluid_intake, collapse=" "), "\n"))
}

methods(class='medicalRecord')
# yay! print shows up

suppressWarnings(print.medicalRecord(patient_record)) # it doesn't like that I have multiple arguments
class(patient_record)
```

2. Write a `medicalRecord` method for the generic function `mean`, which returns averages for pulse, temperature and fluids. Also write a `medicalRecord` method for `print`, which employs some nice formatting, perhaps arranging measurements by date, and `plot`, that generates a composite plot of measurements over time.  Call each function for the medical record created in part 1. (5 points)
```{r echo=TRUE, error=TRUE}
# mean
mean.medicalRecord <- function(record) {
  cat(sprintf("mean pulse: %s\nmean temperature: %s\nmean fluids: %s\n",
              mean(record$pulse), mean(record$temperature), mean(record$fluid_intake), "\n"))
}

suppressWarnings(mean.medicalRecord(patient_record))

# I kinda already did this up above, so here's a modified print method for medicalRecord
# be careful if sorting by date - need to also sort the corresponding measurements
print.medicalRecord.date <- function(record) {
  df <- data.frame(record$date_of_admission, # if there are multiple measurements, arrange by date
                   record$pulse,
                   record$temperature,
                   record$fluid_intake
                 )
  sorted_df <- df[order(df[,1], decreasing=FALSE), ] # sort by the date column, then extract values back out for the medical record
  cat(sprintf("name: %s\ngender: %s\ndate_of_birth: %s\ndate_of_admission: %s\npulse: %s\ntemperature: %s\nfluid_intake: %s", 
              record$name, record$gender, record$date_of_birth, paste(sorted_df[,1], collapse=" "), paste(sorted_df[,2], collapse=" "), paste(sorted_df[,3], collapse=" "), paste(sorted_df[,4], collapse=" "), "\n"))

}

# TS: It's better to summarize date of admission and those three measures using dataframe. See example code below.
print.medicalRecord <- function(x, ...) {
  dateix <- order(as.POSIXct(x$date_of_admission))
  a <- sprintf("Name: %s\nGender: %s\nDate of Birth: %s\n", x$name, x$gender, x$date_of_birth)
  b <- data.frame(x$date_of_admission[dateix], x$pulse[dateix],
                  x$temperature[dateix], x$fluid_intake[dateix])
  names(b) <- c("Admission Date", "Pulse", "Temperature", "Fluid Intake")
  cat(a)
  print(b)
}

suppressWarnings(print.medicalRecord.date(patient_record))

install.packages('gridExtra')
library(ggplot2)
library('gridExtra')

plot.medicalRecord <- function(record) {
  df <- data.frame(record$date_of_admission, # I want to keep this data frame for plotting
                   record$pulse,
                   record$temperature,
                   record$fluid_intake
                 )
  sorted_df <- df[order(df[,1], decreasing=FALSE), ]
  dates=as.Date(sorted_df[,1])
  pulse=sorted_df[,2]
  temp=sorted_df[,3]
  fluids=sorted_df[,4]
  # I want three plots in one window, since each measurement is on a different scale
  # not perfect, but I couldn't combine them because of the different y axes
  a <- ggplot(sorted_df, aes(x=dates, y=pulse)) + geom_line() + geom_point() + scale_x_date(date_labels = "%Y-%m-%d")
  b <- ggplot(sorted_df, aes(x=dates, y=temp)) + geom_line() + geom_point() + scale_x_date(date_labels = "%Y-%m-%d")
  c <- ggplot(sorted_df, aes(x=dates, y=fluids)) + geom_line() + geom_point() + scale_x_date(date_labels = "%Y-%m-%d")
  grid.arrange(a, b, c, nrow=3)
}

plot.medicalRecord(patient_record)
```

3. Create a further class for a cohort (group) of patients, and write methods for `mean` and `print` which, when applied to a cohort, apply mean or print to each patient contained in the cohort. Hint: think of this as a "container" for patients.  Reset the RNG seed to 8 and create a cohort of ten patients, then show the output for `mean` and `print`. (5 points)
```{r echo=TRUE, error=TRUE}
set.seed(8)
cohort <- replicate(10, makePatient(), simplify=FALSE) # 10 patients
class(cohort) <- c('cohortRecord', 'medicalRecord')
inherits(cohort, 'medicalRecord')

#cohort[[1]] # first person in the cohort
#cohort[[1]][1] # name
#cohort[[1]][2] # gender
#cohort[[1]][3] # date_of_birth
#cohort[[1]][4] # date_of_admission
#cohort[[1]][5] # pulse
#cohort[[1]][6] # temperature
#cohort[[1]][7] # fluid_intake

for (i in 1:(length(cohort))) { # cohort is in groups of 7 (per individual)
  names(cohort[[i]]) <- c('name', 'gender', 'date_of_birth', 'date_of_admission', 'pulse', 'temperature', 'fluid_intake')
}

# don't tell on me, pretty sure this would violate HIPAA
mean.cohort <- function(record) { # where record is the overall cohort
  for (i in 1:(length(record))) {
    cat(record[[i]]$name, '\n')
    suppressWarnings(mean.medicalRecord(record[[i]]))
    cat('\n') # skips a line so you can look at each patient's average info a little more easily
  }
}
mean.cohort(cohort)

print.cohort <- function(record) {
  for (i in 1:(length(record))) {
    suppressWarnings(print.medicalRecord.date(record[[i]]))
    cat('\n\n')
  }
}
print.cohort(cohort)
```
